---
title: "Climate Extremes Analysis"
author: "Abigail Sanford, Julia Bickford, Andrea Cornelius"
date: "4/17/2023"
output:
  html_document:
    toc: yes
    code_folding: hide
---

# Read in the Data

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# load pacakges
library(lubridate)
library(tidyverse)
```

```{r readdata}
clim <- read.table("thermal_precip_temp.csv", fill=TRUE, sep=',', header=T) # read in the data

filter(clim, NAME == "DESERT RESORTS REGIONAL AIRPORT, CA US") # Only keep data from the station that has both temp and precip info
  
head(clim) #this lets us look at our data table in the Rmarkdown rather than having to navigate to the data table in the environment

climtidy <- clim %>% 
  select(DATE, PRCP, TMAX, TMIN) %>% # selected the columns I want to keep which include date, precipitation, temperature max, and temperature minimum
  mutate(DATE = ymd(DATE)) %>% # converts the date column into the 'date' class
  mutate(YEAR = year(DATE)) # Create new column with just the year in it
  


```

# How many heat wave days are there per year?

### Heatwave Threshold set to 80F
```{r}
heatwavedays <- climtidy %>% 
  mutate(HEATWAVE = ifelse(clim$TMAX >= 80, 1, 0)) %>% # created a new column that puts a 1 down if temp was 80F or higher, and a 0 for all other temperatures. If different heatwave threshold is selected, change the 80 to the new heatwave threshold
  drop_na() # remove NA values, this removed 12,276 observations...do we need to anything about that
  
  
```

```{r}
# look at data in a chart

heatwavedaystidy <- heatwavedays %>% 
  select(YEAR, HEATWAVE) %>% # pull out only year and heatwave columns
  group_by(YEAR) %>% # have r recognize that the data can be organized into years
  summarise(across(c(HEATWAVE), # add up the number of days a heat wave has occurred in each year
                   list(sum = sum)))

heatwavedaystidy # view table
```

```{r}
# visualize heatwave days in a graph

ggplot(heatwavedaystidy, aes(x = YEAR, y = HEATWAVE_sum)) +
  geom_point(color = "darkorange4", size = 2) +
  geom_text(mapping = aes (label = HEATWAVE_sum), 
            position = position_dodge(width = 0.9), 
            cex = 3, 
            vjust = -0.9) +
  geom_smooth(method = 'lm', se = FALSE, color = "dimgrey") +
  labs(y = "Number of Heatwave Days (Threshold 80F)",
       x = "Year",
       title = "Annual Amount of Heatwave Days in Thermal California (1998-2022)") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black")) + 
    ylim(0, 500) +
  scale_x_continuous(breaks = c(1945, 1955, 1965, 1975, 1985, 1995, 2005, 2015))


```
#### Brief Analysis
There is a slight increasing trend in the amount of heatwave days per year, given that our heat wave threshold is 80F

### Heatwave Threshold set to 90F
```{r}
heatwavedays90 <- climtidy %>% 
  mutate(HEATWAVE = ifelse(clim$TMAX >= 90, 1, 0)) %>% # created a new column that puts a 1 down if temp was 90F or higher, and a 0 for all other temperatures. If different heatwave threshold is selected, change the 90 to the new heatwave threshold
  drop_na() # remove NA values

```

```{r}
# look at data in a chart

heatwavedaystidy90 <- heatwavedays90 %>% 
  select(YEAR, HEATWAVE) %>% # pull out only year and heatwave columns
  group_by(YEAR) %>% # have r recognize that the data can be organized into years
  summarise(across(c(HEATWAVE), # add up the number of days a heat wave has occurred in each year
                   list(sum = sum)))

heatwavedaystidy90 # view table
```


```{r}
# visualize heatwave days in a graph

ggplot(heatwavedaystidy90, aes(x = YEAR, y = HEATWAVE_sum)) +
  geom_point(color = "darkorange4", size = 4) +
  geom_text(mapping = aes (label = HEATWAVE_sum), 
            position = position_dodge(width = 0.9), 
            cex = 3, 
            vjust = -0.9) +
  geom_smooth(method = 'lm', se = FALSE, color = "dimgrey") +
  labs(y = "Number of Heatwave Days (Threshold 90F)",
       x = "Year",
       title = "Annual Amount of Heatwave Days in Thermal California (1998-2022)") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black")) + 
    ylim(0, 365) +
  scale_x_continuous(breaks = c(1998, 2002, 2006, 2010, 2014, 2018, 2022))


```


### Heatwave Threshold set to 100F
```{r}
heatwavedays100 <- climtidy %>% 
  mutate(HEATWAVE = ifelse(clim$TMAX >= 100, 1, 0)) %>% # created a new column that puts a 1 down if temp was 100F or higher, and a 0 for all other temperatures. If different heatwave threshold is selected, change the 100 to the new heatwave threshold
  drop_na() # remove NA values

```

```{r}
# look at data in a chart

heatwavedaystidy100 <- heatwavedays100 %>% 
  select(YEAR, HEATWAVE) %>% # pull out only year and heatwave columns
  group_by(YEAR) %>% # have r recognize that the data can be organized into years
  summarise(across(c(HEATWAVE), # add up the number of days a heat wave has occurred in each year
                   list(sum = sum)))

heatwavedaystidy100 # view table
```


```{r}
# visualize heatwave days in a graph

ggplot(heatwavedaystidy100, aes(x = YEAR, y = HEATWAVE_sum)) +
  geom_point(color = "darkorange4", size = 4) +
  geom_text(mapping = aes (label = HEATWAVE_sum), 
            position = position_dodge(width = 0.9), 
            cex = 3, 
            vjust = -0.9) +
  geom_smooth(method = 'lm', se = FALSE, color = "dimgrey") +
  labs(y = "Number of Heatwave Days (Threshold 100F)",
       x = "Year",
       title = "Annual Amount of Heatwave Days in Thermal California (1998-2022)") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black")) + 
    ylim(0, 365) +
  scale_x_continuous(breaks = c(1998, 2002, 2006, 2010, 2014, 2018, 2022))


```

### Heatwave Threshold set to 110F
```{r}
heatwavedays110 <- climtidy %>% 
  mutate(HEATWAVE = ifelse(clim$TMAX >= 110, 1, 0)) %>% # created a new column that puts a 1 down if temp was 110F or higher, and a 0 for all other temperatures. If different heatwave threshold is selected, change the 110 to the new heatwave threshold
  drop_na() # remove NA values

```

```{r}
# look at data in a chart

heatwavedaystidy110 <- heatwavedays110 %>% 
  select(YEAR, HEATWAVE) %>% # pull out only year and heatwave columns
  group_by(YEAR) %>% # have r recognize that the data can be organized into years
  summarise(across(c(HEATWAVE), # add up the number of days a heat wave has occurred in each year
                   list(sum = sum)))

heatwavedaystidy110 # view table
```


```{r}
# visualize heatwave days in a graph

ggplot(heatwavedaystidy110, aes(x = YEAR, y = HEATWAVE_sum)) +
  geom_point(color = "darkorange4", size = 4) +
  geom_text(mapping = aes (label = HEATWAVE_sum), 
            position = position_dodge(width = 0.9), 
            cex = 3, 
            vjust = -0.9) +
  geom_smooth(method = 'lm', se = FALSE, color = "dimgrey") +
  labs(y = "Number of Heatwave Days (Threshold 110F)",
       x = "Year",
       title = "Annual Amount of Heatwave Days in Thermal California (1998-2022)") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black")) + 
    ylim(0, 365) +
  scale_x_continuous(breaks = c(1998, 2002, 2006, 2010, 2014, 2018, 2022))


```

### Heatwave Threshold set to 120F
```{r}
heatwavedays120 <- climtidy %>% 
  mutate(HEATWAVE = ifelse(clim$TMAX >= 120, 1, 0)) %>% # created a new column that puts a 1 down if temp was 80F or higher, and a 0 for all other temperatures. If different heatwave threshold is selected, change the 80 to the new heatwave threshold
  drop_na() # remove NA values

```

```{r}
# look at data in a chart

heatwavedaystidy120 <- heatwavedays120 %>% 
  select(YEAR, HEATWAVE) %>% # pull out only year and heatwave columns
  group_by(YEAR) %>% # have r recognize that the data can be organized into years
  summarise(across(c(HEATWAVE), # add up the number of days a heat wave has occurred in each year
                   list(sum = sum)))

heatwavedaystidy120 # view table
```


```{r}
# visualize heatwave days in a graph

ggplot(heatwavedaystidy120, aes(x = YEAR, y = HEATWAVE_sum)) +
  geom_point(color = "darkorange4", size = 4) +
  geom_text(mapping = aes (label = HEATWAVE_sum), 
            position = position_dodge(width = 0.9), 
            cex = 3, 
            vjust = -0.9) +
  geom_smooth(method = 'lm', se = FALSE, color = "dimgrey") +
  labs(y = "Number of Heatwave Days (Threshold 120F)",
       x = "Year",
       title = "Annual Amount of Heatwave Days in Thermal California (1998-2022)") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black")) + 
    ylim(0, 365) +
  scale_x_continuous(breaks = c(1998, 2002, 2006, 2010, 2014, 2018, 2022))


```

# How long do the heatwaves last, when there is a heat wave?



# How long does summer last each year?

# Has precipitations been going down as well?



### Extremes metrics: number of freezing days

We'll look at a couple different examples of metrics for extremes. First, let's take the number of freezing days: this is defined as the number of days (per year) with minimum temperatures below freezing. Here the data is in degrees Fahrenheit, so we'll select all days where the `TMIN` field in our data table is below 32. Adjusting to be `Tmax` field in our data table is higher than 80

The code block below assigns a new field to the `clim` table, called "freeze": this is defined using the `ifelse` statement. Ifelse is a handy R function which returns one value if a logical statement passed to the function is true, and a different one if the statement is false. In this case, we're telling R to evaluate the statement "is the minimum temperature below 32F?": if yes, clim$freeze is assigned a value of 1, and if no, it's assigned a value of zero.

Once the ifelse statement has been completed, the next thing the code below does is make use of the `group_by` and `summarize` functions to count the number of freezing days! This is done in two steps:

* Grouping the data by year (`clim %>% group_by(year(date))`)
* Then passing the grouped data to `summarize` and totaling up the number of times during each year when the minimum temperature is below 32 (recall that the `freeze` field is 1 each time this is true! that makes it easy to count days, simply by adding up all the ones)

The times that go along with the (now yearly!) number of freezing days are calculated by finding all the unique years in the dataset (`unique(year(date))`) and assigned to another new field in the data table called `dt`.

Finally, the number of freezing days is plotted using ggplot, as we did for the trends demo.


```{r other metrics, echo=TRUE}

clim$heatwave = ifelse(clim$TMAX >= 80, 1, 0)
clim.wfrdata = clim %>% group_by(year(date)) %>% summarize(ndayfr=sum(heatwave))
clim.wfrdata$dt = unique(year(date))

ggplot(clim.wfrdata, aes(dt, ndayfr))+geom_point()+labs(y="Number of Heatwave Days")
```

Note that the trends in mean-state and extreme metrics often go together: from last time, recall that there was a trend toward decreasing minimum daily temperature. How does that affect the frequency of occurrence of freezing days?


```{r other metrics, echo=TRUE}

format(as.Date(clim$DATE, format="%m/%d/%Y"),"%Y")


clim$heatwavedays = ifelse(clim$TMAX >= 80, 1, 0)



clim.heatwavedata = clim %>% group_by(year(date)) %>% summarize(ndayfr=sum(heatwavedays))


ggplot(clim.wfrdata, aes(date, ndayfr))+geom_point()+labs(y="Number of Heatwave Days")
```



### Extremes metrics: hottest day of the year

Looking at the warm end of the temperature distribution, we can now look for trends in the temperature during the hottest day of the year.

This is accomplished using a similar methodology as we used for the freezing days above; again, the group_by function groups the data according to year, then the temperature associated with the hottest day is calculated by applying the `max` function to the `TMAX` field in the data table.

```{r hottest days, echo=TRUE}
clim.pk = clim %>% group_by(year(date)) %>% summarize(hotday=max(TMAX))
clim.pk$dt = unique(year(date))

ggplot(clim.pk, aes(dt, hotday))+geom_point()+labs(y="Hottest Day in the Year")
```

### Extremes metrics: return periods

In lecture, we discussed the concept of return periods and return levels. Here is how you calculate these things with real data!

We'll specify some reasonable threshold for daily precipitation: say, 1 inch/day. To calculate the return period associated with that threshold, what we do is the following:
* Use `ifelse` to create a field called "flood" that is 1 when precipitation is above the threshold, and 0 otherwise
* Store the total number of years by first identifying the total number of unique years, then calculating the length of the array where those unique years are listed

The return period is then the number of years (plus one to include the starting year) divided by the number of times the threshold exceedance has actually occurred. (Note: we have to use the `na.rm=TRUE` flag to prevent NA entries from seeming like they are contributing to the count of threshold exceedances!)

``` {r return period}
clim$flood = ifelse(clim$PRCP >= 1, 1, 0)
nyrs=length(unique(year(date)))

retper=(nyrs+1)/sum(clim$flood,na.rm=TRUE)

retper
```

Another note: return *levels* are somewhat more complicated to calculate, since they require fitting the underlying distribution of the data... there are lots of resources out there if you need them though!
